 In general, R requires dataset to be loaded into memory. A notable feature of the raster package is that it can work with   raster datasets that are stored on disk and are too large to be loaded into memory
 (RAM). The package is built around a number of 'S4' classes of which the RasterLayer,
 RasterBrick, and RasterStack classes are the most important.
 A RasterLayer object represents single-layer (variable) raster data. A RasterLayer
 object always stores a number of fundamental parameters that describe it. These
 include the number of columns and rows, the coordinates of its spatial extent
 ('bounding box'), and the coordinate reference system (the 'map projection').
 In addition, a RasterLayer can store information about the file in which the
 raster cell values are stored (if there is such a file). A RasterLayer can also
 hold the raster cell values in memory.
 
 The raster package has two classes for multi-layer data the RasterStack and the RasterBrick.

Unsigned Integer vs Signed Integer
an unsigned integer is always non-negative numbers. But a signed integer can store negative values.
With an 8-bit unsigned raster, valid values are from 0 to 255. This means that an 8-bit raster has 256 values in total.


```{r}
# load the raster, sp, and rgdal packages
install.packages("pacman")

```

    Installing package into ‘/home/nbuser/R’
    (as ‘lib’ is unspecified)



```{r}
install.packages("rgdal")
```

    Installing package into ‘/home/nbuser/R’
    (as ‘lib’ is unspecified)
    Warning message in install.packages("rgdal"):
    “installation of package ‘rgdal’ had non-zero exit status”


```{r}
library(pacman)
p_load(rgdal)
p_load(raster)
p_load(sp)
p_load(sf)
p_load(fasterize)
```
download gridded population data

```{r}
dir.create("rasterdata")
setwd(file.path(getwd(),"rasterdata"))
getwd()
rf1<-"NPL_ppp_2020_v2.tif"
rf2<-"NPL_pph_2020_v2.tif"

if(!file.exists(rf1)){
file.path1 <-"ftp://ftp.worldpop.org.uk/GIS/Population/Individual_countries/NPL/Nepal_100m_Population/NPL_ppp_2020_v2.tif"
    download.file(file.path1, rf1,  mode = "wb")
        }
if(!file.exists(rf2)){
file.path2 <-"ftp://ftp.worldpop.org.uk/GIS/Population/Individual_countries/NPL/Nepal_100m_Population/NPL_pph_2020_v2.tif"
    download.file(file.path2, rf2,  mode = "wb")
        }
```


```{r}
setwd(file.path(getwd(),"rasterdata"))
rf1<-"NPL_ppp_2020_v2.tif"
rf2<-"NPL_pph_2020_v2.tif"
file.exists(rf1)
file.exists(rf2)
```


TRUE



TRUE



```{r}
## ----load-raster---------------------------------------------------------
# load raster in an R object called PD (population density)

```



```{r}
PD<-raster(rf1)
# look at the raster attributes. 
PD

# calculate and save the min and max values of the raster to the raster object
PD <- setMinMax(PD)

# view raster attributes
PD

#Get min and max cell values from raster
#NOTE: this code may fail if the raster is too large
cellStats(PD, min)
cellStats(PD, max)
```

view coordinate reference system

```{r}
## ----crs-----------------------------------------------------------------

PD@crs
```

view raster extent
```{r}
## ----view-extent---------------------------------------------------------

PD@extent

```


```{r}
## ----crop raster-------------------------------------------------
## 
plot(PD)
## #Define the extent of the crop by clicking on the plot
cropbox1 <- drawExtent()
## #crop the raster, then plot the new cropped raster
PDcrop1 <- crop(PD, cropbox1)
## 
## #plot the cropped extent
plot(PDcrop1)
## 
```


```{r}
## ----crop Kathmandu area-----------------------------------------

#download Nepal shapefile
nepal_shp<-getData('GADM', country="NEPAL", level=3)
plot(nepal_shp)
#select Kathmandu region
Kathmandu_shp<-nepal_shp[nepal_shp$NAME_3 %in% c("Kathmandu", "Bhaktapur", "Lalitpur"),]

plot(Kathmandu_shp)
invisible(text(coordinates(Kathmandu_shp), labels=as.character(Kathmandu_shp$NAME_3)))

#Crop Kathmandu region
PD_kath <- crop(PD, Kathmandu_shp)
```


```{r}
## ----histogram-----------------------------------------------------------

# the distribution of values in the raster
hist(PD_kath, main="Distribution of population counts", col= "purple")

```


```{r}
## ----plot-raster---------------------------------------------------------
# add a color map with 5 colors
col=topo.colors(5)
# specify the colors using color-ramp 
colr <- colorRampPalette(c("navyblue", "steelblue", "limegreen", "yellow", "#FEFEFE"))(255)
# add breaks to the colormap (6 breaks = 5 segments)
brk <- c(0, 5, 10, 35, 100, 1100)

# create a plot of our raster
image(PD_kath, col = col)
```


```{r}
# specify the range of values that you want to plot in the PD
# just plot pixels between 0 and 100
image(PD_kath, zlim=c(0,100), col = col, main = "Population Distribution")
#overlay boundary
plot(Kathmandu_shp, add=T, border='white')

```


```{r}
#Plot cropped area
plot(PD_kath, main="Population Distribution - Kathmandu", col=colr)
#overlay boundary
plot(Kathmandu_shp, add=T, col=rgb(red=0, blue=0, green=0, alpha=0.1))

```


```{r}
## ----plot-with-breaks----------------------------------------------------
plot(PD_kath, col=col, breaks=brk, main="Counts with more breaks", legend = FALSE)

```


```{r}
#project raster to 100 meters resolution

#define a projection: https://proj4.org/usage/projections.html
 prj<-"+proj=tmerc +lat_0=0 +lon_0=84 +k=0.9996 +x_0=500000 +y_0=0 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"
### One single line is sufficient to project any raster to any projection
 PD_kath_prj<-projectRaster(PD_kath, crs=crs(prj), res = 100)
```


```{r}
#multilayer raster example
 # Using the previously generated RasterLayer object
 r<-PD_kath
 # Let's first put some values in the cells of the layer
 r[] <- rnorm(n=ncell(r))
 # Create a RasterStack object with 3 layers
 s <- stack(x=c(r, r*2, r^2))
 # The exact same procedure works for creating a RasterBrick
 b <- brick(x=c(r, r*2, r))
 # Let's look at the properties of of one of these two objects
 b
```


```{r}
 #load raster stack from disk
 listras<-list.files(pattern = ".tif$", full.names = T)
 #listras<-listras[stringr::str_detect(listras, 'ppp')]
 
 stackras<-stack(listras)
 plot(stackras)
 
```


```{r}
 #vector to raster conversion
 temp_ras <- PD
 temp_ras[]<-NA
 
#load census level 4 data
 #pop_sf<-st_read("./NPL_level4/ADMINPOP.shp")
 pop_sf<-getData('GADM', country="NEPAL", level=4)
 pop_sf$ID_4<-1:nrow(pop_sf)
 gridz <- rasterize(pop_sf, temp_ras, field = "ID_4");
 #much faster using fasterize but requires sf format data
 pop_sf<-st_as_sf(pop_sf)
 gridz1 <- fasterize::fasterize(pop_sf, temp_ras, field = "ID_4");
 compareRaster(gridz, gridz1)
 
```


```{r}
#Map Algebra Types
 #Map algebra can be defined as local, focal, zonal and global operations.
 
 #local operation on raster such change the value if individual cell based on some transformation
 logPD<-log10(PD)
```


```{r}
#calculate zonal statistics based on zone
 zonal_grid <-zonal(PD, gridz, fun = "sum", na.rm = TRUE);
head(zonal_grid)
```


```{r}
#focal operation  such as moving average of 3 x 3 cells
 focal_grid <- focal(PD, w=matrix(1/9, nc=3, nr=3))
 head(focal_grid)
```


```{r}
#global operation such as global sum
 cellStats(PD, sum)
 cellStats(PD, sd)
```
